A Bayesian method for latitudinal temperature gradients in the early Eocene

### 2023-02-08 manuscript/manuscript.Rmd
update discussion

Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.

Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1 degree latitudinal bins. The weights were set to the proportion of global surface area in each latitudinal bin, i.e. decreasing with increasing latitude as:
$$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$$
where $\alpha1$ is the upper, and $\alpha2$ is the lower latitudinal boundary of bin $i$.

@@

Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global average temperatures. The modern, global mean sea surface temperature estimate is `r motemp_global$"50%"` degC when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern global mean sea surface temperature is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"` degC.

@@

The global early Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).

@@

Modelling the latitudinal temperature gradient from the equator to the poles facilitates reconstruction of the global mean temperature.

>>>>

Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.  
Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1 degree latitudinal bins. The weights were set to the proportion of global surface area in each latitudinal bin, i.e. decreasing with increasing latitude as:
$$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$$
where $\alpha1$ is the upper, and $\alpha2$ is the lower latitudinal boundary of bin $i$.

@@

Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST). The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"` degC when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"` degC.

@@

The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC,  `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)` degC  higher than the modern.

@@

Palaeoclimate reconstructions are often summarised as global mean surface temperatures, to provide a standardised metric for characterising the state of the Earth's climate [@Royer2004;@Inglis2020]. The calculation of global mean surface temperatures from sparse proxy data is susceptible to bias [@Jones2022]. By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record through inter- or extrapolation, eliminating the problem that specific climate zones may dominate the proxy record. Intersecting the modeled temperature gradient at narrow latitudinal intervals and accounting for the varying sizes of latitudinal bands by weighting provides unbiased GMST estimates. Recent deep-time GMST reconstructions are sometimes accounting for different sampling by calculating zonal averages, but assum [@Inglis2020]. We anticipate that applying this method to intervals with a small or biased sample of temperature proxies will change Phanerozoic, proxy-based temperature curves [@Royer2004;@Veizer2015;@Jones2022].

Estimating the GMST without bias is achieved by intersecting the modeled temperature gradient at narrow latitudinal intervals and adjusting for the differing sizes of latitudinal bands through weighting.

Estimating the GMST from sparse proxies is prone to bias [@Jones2022], and

Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes, which allows to estimate GMST even

Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes, which can be used to improve estimates of the global mean temperature. Even if the underlying proxy data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques. The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.

popularity of reporting global mean temperature in paleoclimate studies is due to its utility as a comprehensive and standardized metric for characterizing and comparing the state of the Earth's climate and evaluating climate models.

Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes. This information can be used to improve estimates of the global mean temperature. Even if the underlying proxy data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques. The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.

Modeling the latitudinal temperature gradient provides a comprehensive representation of the temperature distribution across different latitudes. This information can be used to make more informed estimates of the global mean temperature. Even if the data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques. The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.
